Article Title

### 2013-07-28 sections/Introduction.tex
expanded the section on the theory of regularization.

In this study, we pursue two goals: 
\begin{enumerate}[\qquad A.\;\;]
	\item Devise better estimators of neural covariance matrices in recordings from large populations of neurons.
	\item Aid interpretation of population activity  by isolating the essential structure of the neural covariance matrix.
\end{enumerate}

The two goals are inextricably connected as they both benefit from reduced, simplified approximations of the covariance matrix.

In estimation theory, the overall error of an estimate can be reduced by \emph{regularization}, which is the deliberate biasing of the solution toward a simplified, low-dimensional (`sparse') \emph{target estimate} \citep{Schafer:2005,Bickel:2006}.
Regularized estimators allow reaching a favorable tradeoff between bias and variance in order to minimize the total error.
Strikingly, some improvement can be produced even when the target estimate is chosen arbitrarily, which is sometimes described as ``Stein's phenomenon'' after its discoverer \cite{Stein:1956}.   However, when a sparse target estimate does happen to match the dominant features of the true covariance matrix, it will induce relatively little bias while retaining low variance.  For example, some covariance estimators have been motivated by a specific model of the underlying process such as single-factor models of the stock market, for example \citep{Ledoit:2003}. 
Since the true covariance structure is not known in practice, the performance of estimators is evaluated by how well their performance generalized to new data, outside the training sample by cross validation, for example.

The above consideration served as the basis of our approach.  We proposed several alternative candidate estimators of the neural covriance matrix in a given circuit, each motivated by a different hypothesis for the origin of neural correlations.  When a consistent winner emerges from cross-validated performance comparisons, the sparse estimate used in its regularization can be interpreted to reflect a simplified correlation structure supported by the emprical data.

>>>>

With this study, we pursue two goals: 
\begin{enumerate}[\qquad A.\;\;]
	\item Devise better estimators of neural covariance matrices in recordings from large populations of neurons.
	\item Aid interpretation of population activity  by isolating the essential structure of the neural covariance matrix.
\end{enumerate}

To select the best covariance matrix estimate for a specific neural circuit, we evaluate the performance of four estimators using different regularization schemes, each motivated by  a different hypothesis for the origin of neural correlations. \emph{Regularization} is the deliberate biasing the solution toward a simplified, low-dimensional (`sparse') \emph{target estimate} \citep{Schafer:2005,Bickel:2006}.
Regularized estimators allow reaching a favorable tradeoff between bias and variance in order to minimize the total error.
Strikingly, some improvement can be produced even when the target estimate is chosen arbitrarily, which is sometimes described as ``Stein's phenomenon'' after its discoverer \cite{Stein:1956}.   However, when a sparse target estimate does happen to match the dominant features of the true covariance matrix, it will induce relatively little bias while retaining low variance.  For example, some covariance estimators have been motivated by a specific model of the underlying process such as single-factor models of the stock market, for example \citep{Ledoit:2003}. 
Since the true covariance structure is not known in practice, the performance of estimators is evaluated by how well their performance generalized to new data, outside the training sample by cross validation, for example.

### 2013-07-28 sections/Theory.tex
improved introduction, added figure captions.
expanded the section on the theory of regularization.

\subsection{Problem setup}
Let $x_1,\ldots,x_n$ denote a sample of observations of instantaneous spiking rates of $p$ neurons in time bins $i=1,\ldots,n$.  Each obseravation $x_i$ is a $p\times 1$ vector of spike counts.  We assume that observations independent, identically distributed, i.e.\;all observations are drawn from a common true distribution $x_1,\ldots,x_n \sim f$.  However, this assumptions can be relaxed and corrected for as discussed in this section. We make no assumption about the exact form of distribution $f$.

Then the true  covariance matrix $\Sigma$ is defined as
\begin{equation}
\Sigma =\mathbb E_z[ (z-\mathbb E_{z\sim f} [z])(z-\mathbb E_z [z])^\T]
\end{equation}
where $\mathbb E_z[g(z)]$ denotes the expected value of $g(z)$ when $z$ is distributed according to $f(z)$.

The purpose of the covariance estimation procedure is to obtain an estimate $\hat \Sigma$ that is as close as possible to $\Sigma$ given the finite sample  $x_1,\ldots,x_n$. We define a scalar \emph{loss function} $\mathcal L\left(\hat \Sigma, \Sigma\right)$ to measure the distance between $\hat \Sigma$ and $\Sigma$.  The loss function must have minimum at $\hat \Sigma=\Sigma$.

In particular, we consider the sum-of-squared-error loss
\begin{equation}
\mathcal L_e(\hat \Sigma, \Sigma) =  \| \hat \Sigma - \Sigma\|^2_F 
\end{equation}
where $\| A \|_F = \sqrt{\Tr(AA^\T)}$ denotes the Frobenius matrix norm; $\Tr(\cdot)$ denotes the matrix trace.

We also consider the normal-theory-log-likelihood loss
\begin{equation}
\mathcal L_g(\hat \Sigma,\Sigma) = \ln |\hat \Sigma|  + \Tr(\hat \Sigma^{-1}\Sigma)
\end{equation}
where $|A|$ denotes the determinant of matrix $A$.

Note that both $\mathcal L_e(\hat\Sigma,\Sigma)$ and $\mathcal L_g(\hat\Sigma,\Sigma)$ attain their minima when $\hat\Sigma = \Sigma$.   $\mathcal L_g$ is derived as the negative log likelihood function (scaled and shifted) when $f$ is a multivariate normal distribution.  However, using either of these loss functions does not amount to assuming a specific form of $f$. Other loss functions have been defended in statistical literature \citep{James:1961,Fan:2008}.

The loss function can be estimated based on another estimate $\hat\Sigma_Y$ of the covariance matrix obtained from an independent \emph{validation} sample $Y = (y_1,\ldots,y_m)\sim f^m$.

Both $\mathcal L_e(\hat\Sigma,\Sigma)$ and $\mathcal L_g(\hat\Sigma,\Sigma)$ are linear with respect to $\Sigma$ in the sense that
\begin{equation}
\mathcal L\left(\hat \Sigma,\alpha S_1 + (1-\alpha)S_2\right) 
\equiv 
\alpha\mathcal L(\hat\Sigma,S_1) + (1-\alpha)\mathcal L(\hat\Sigma,S_2)
\end{equation}
If we ensure that  $\hat\Sigma_Y$ is unbiased, i.e.\;$\mathbb E_Y\left[\hat \Sigma_Y\right]=\Sigma$, then this linearity implies that 
\begin{equation}
\mathbb E_Y \left[\mathcal L(\hat\Sigma,\hat\Sigma_Y)\right]
=
\mathcal L\left(\hat\Sigma,\mathbb E_Y \left[\hat\Sigma_Y\right]\right)
= 
\mathcal L(\hat\Sigma,\Sigma)
\end{equation}

In other words, the empirical loss $\mathcal L(\hat\Sigma,\hat\Sigma_Y)$ is an ubiased estimator of true loss $\mathcal L(\hat\Sigma,\Sigma)$ provided that $\hat\Sigma_Y$ is unbiased and $\hat\Sigma$ and $\hat\Sigma_Y$ are estimated from independent samples.  Note that $\hat\Sigma$ does not need to be unbiased.

In practice, recording a separate independent sample may not be sensible.  Instead, the recorded data a split into a training sample and validation sample.  One popular approach is $K$-fold validation, in which the data are split into $K$ roughly equal-sized subsets.  Each subset then successively serves as the validation  sample while the rest of the data make up the training sample. This procedure yields $k$ estimates of the loss, although not quite independent.

>>>>

\subsection{Covariance estimation}

\subsubsection{Problem setup}

Let $x_i \in \mathcal X, i=1\ldots,x_n$ denote a sample of consequtive observations of population activity of $p$ neurons in time bins $i$.  
The activity of each neuron is the real-valued firing rate, thus  $\mathcal X \in \mathbb R^{p\times 1}$.  
We assume that observations are drawn from some true cumulative probability distribution $F: \mathcal X \mapsto R$. Then let $\hat F_n$ denote the empirical distribution function obtained from the observed sample, which itself is a random variable.
\begin{equation}
\hat F_n(x) = \sum\limits_{i=1}^n \mathbf{1}(x > x_i)
\end{equation}
where $\mathbf 1(x_i \le x)$ is the indicator function which equals 1 when all elements of $x$ are greater than the corresponding elements of $x_i$ and 0 otherwise.

The observations need not be independently and identically sampled from $F$. The only assumption is that the data generating process is ergodic, \emph{i.e.}\;that $\max_x \left[\hat F_n(x) - F(x)\right] \to 0$ as the sample size $n$ increases. Then each $x_i$ can be thought to be drawn from $F$ when not conditioned on nearby observations.

The true covariance matrix $\Sigma \in \Theta$ is defined as a function of the true probability distribution
\begin{equation}
\Sigma(F) = \int\limits_{z\in\mathcal X} (z - \bar x)(z - \bar x)^\T \dif F(z)
\quad\mbox{where}\;
\bar x = \int\limits_{z\in\mathcal X} z \dif F(z)
\end{equation}
The \emph{plug-in estimator} is obtained by plugging the empirical distribution in place of the true distribution
\begin{equation}
\begin{split}
\Sigma(\hat F_n) = \int\limits_{z\in\mathcal X} (z - \bar x)(z - \bar x)^\T \dif \hat F_n(z) 
\equiv \frac  1 n\sum\limits_{i=1}^n (x_i-\bar x)(x_i - \bar x)
\\
\mbox{where}\;\bar x = 
\int\limits_{z\in \mathcal X}z \dif \hat F_n(z) \equiv \frac 1 n\sum\limits_{i=1}^n x_i
\end{split}
\end{equation}
The plugin estimator is known to be biased toward zero because the estimate of the mean $\bar x$ absorbs a fraction of the variance.

The usual unbiased estimate, the sample covariance matrix, rescales the plug-in estimator compensates for the bias
\begin{equation}
\hat\Sigma(\hat F_n) = \frac n {n - d} \Sigma(\hat F_n) \equiv \frac 1 {n-d} \sum\limits_{i=1}^n (x_i - \bar x) (x_i - \bar x)^\T
\end{equation}
When observations are independent, then $d=1$.  But when nearby samples are correlated, covariances become underestimated. Then we can use $d=cn$ where $c>1$ is called the \emph{overdispersion coefficient}, which may be estimated from data.

\subsubsection{Definition of regularization}
Now let's define a sequence of approximate biased estimates $\hat\Sigma_\gamma(\hat F_n)$ indexed by the \emph{regularization intensity}  $\gamma$ such that $\hat\Sigma_{\gamma=0}(\hat F_n)\equiv \hat\Sigma(\hat F_n)$, which become increasingly biased as $\gamma$ increases.

To evaluate the estimate, let's define the \emph{loss function} $\mathcal L : \Theta \times \Theta \mapsto \mathbb R$, which is minimized when its two arguments are equal.  Then \emph{excess loss} $\ell(\hat S, S) = \mathcal L(\hat S, S) - \mathcal L(S,S)$ assumes zero at the minimum.

The overall error, known as \emph{estimator risk} is expressed as
\begin{equation}
r_\gamma = \mathbb E_F\left[ \ell\left(\Sigma_\gamma(P_n),\Sigma\right) \right]
\end{equation}

The expected value of the approximate estimate 
$\Sigma_\gamma^{(n)} = \mathbb E_F\left[\hat \Sigma_\gamma(\hat F_n)\right]$ 
is biased with the amout of bias or \emph{approximaton error} expressed as  
\begin{equation}
b_\gamma^2 = \ell \left( \Sigma_\gamma^{(n)},\Sigma\right)
\end{equation}

Then the \emph{estimation error} 
\begin{equation}
\varepsilon_\gamma^2 = \mathbb E_F \left[ \ell\left(\hat \Sigma_\gamma(\hat F_n), \Sigma_\gamma^{(n)}\right) \right]
\end{equation}

When the loss function is the sum of squared error (Euclidean norm squared), which, for matricies, is called the Frobenius norm, $\mathcal L(\hat S,S) = \mathcal L_e(\hat S,S) = \|\hat S-S\|_F^2 = \Tr\left((\hat S-S)(\hat S-S)^\T\right)$, the estimation error is called \emph{variance} and the overall error is expressed as the the sum 
\begin{equation}
r_\gamma =  b_\gamma^2 + \varepsilon_\gamma^2
\end{equation}

With other loss functions, the overall error is also an increasing function of both kinds of errors, but not necessarily a simple sum.

One other popular loss function for covariance estimation arises from the theory of normal distributions.  Under the assumption of normality  with zero mean, the log likelihood of the covariance matrix $\Sigma$ given the independent sample $z_1,\ldots,z_m$, with $S = \tfrac 1 m \sum\limits_{i=1}^m z_i z_i^\T$,  
\begin{equation}
L\left(\Sigma \mid S\right) = -\tfrac m 2 \ln(2\pi) - \tfrac m 2 \ln \det \Sigma - \tfrac m 2 \Tr(\Sigma^{-1} S)
\end{equation}
Then the Gaussian loss function $\mathcal L_g$ is constructed by rescaling the normal log likelihood and dropping the constant term
\begin{equation}
\begin{split}
\mathcal L_g(\hat S,S) = &-\frac 2 m L\left(S \mid \hat S \right) - \ln(2\pi) 
\\ 
\equiv & \ln \det \hat S + \Tr(\hat S^{-1} S) 
\end{split}
\end{equation}
The corresponding excess loss 
\begin{equation}
\ell_g(\hat S,S) = \mathcal L_g(\hat S,S) - \mathcal L_g(S,S) 
= -\ln \det \left(\hat S^{-1} S\right) + \Tr\left(\hat S^{-1}S\right) - p
\end{equation}
is known as \emph{entropy loss} \citep{James:1961}.

Despite the fact that entropy loss is inspired by the normal theory, the choice of a loss function is not equivalent to assuming a specific form of $F$. The loss function only measures the distance between the estimate and the true value of a parameter of the true distribution but not the distance between distributions themselves. Minimizing the loss function improves the estimate under any true distribution.

The task of regulazation comprises the choice of the \emph{regularization path} $\hat \Sigma_\gamma(\hat F_n)$ and the optimal location $\hat\gamma = \gamma(\hat F_n)$ on that path in oder to minimize the risk $r_\gamma$.  
By definition $\hat\Sigma_{\gamma=0}(\hat F_n)$ is an unbiased estimator with $b_{\gamma=0}^2 = 0$.

Both $\mathcal L_e(\hat\Sigma,\Sigma)$ and $\mathcal L_g(\hat\Sigma,\Sigma)$ are linear with respect to $\Sigma$ in the sense that
\begin{equation}
\mathcal L\left(\hat \Sigma,\alpha S_1 + (1-\alpha)S_2\right) 
\equiv 
\alpha\mathcal L(\hat\Sigma,S_1) + (1-\alpha)\mathcal L(\hat\Sigma,S_2)
\end{equation}
If we ensure that  $\hat\Sigma_Y$ is unbiased, i.e.\;$\mathbb E_Y\left[\hat \Sigma_Y\right]=\Sigma$, then this linearity implies that 
\begin{equation}
\mathbb E_Y \left[\mathcal L(\hat\Sigma,\hat\Sigma_Y)\right]
=
\mathcal L\left(\hat\Sigma,\mathbb E_Y \left[\hat\Sigma_Y\right]\right)
= 
\mathcal L(\hat\Sigma,\Sigma)
\end{equation}
In other words, the empirical loss $\mathcal L(\hat\Sigma,\hat\Sigma_Y)$ is an ubiased estimator of true loss $\mathcal L(\hat\Sigma,\Sigma)$ provided that $\hat\Sigma_Y$ is unbiased and $\hat\Sigma$ and $\hat\Sigma_Y$ are estimated from independent samples.  Note that $\hat\Sigma$ does not need to be unbiased.

In practice, recording a separate independent sample may not be sensible.  Instead, the recorded data a split into a training sample and validation sample.  One popular approach is $K$-fold \emph{cross-validation}, in which the data are split into $K$ roughly equal-sized subsets.  Each subset then successively serves as the validation  sample while the rest of the data make up the training sample. This procedure yields $k$ estimates of the loss, although not quite independent.
